---
title: "rmd_demo"
author: "Raphael Morsomme"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

This document contains a few tips for working with RMD files. The source file has been sent by email.

# Level-1 title (often too large)

## Level-2 title (typically better than level-1)

You always want to start by loading the packages you will need.

```{r, message = FALSE}
library(tidyverse)
library(broom)
```

The next step typically consists in downloading the data

```{r}
d <- read_csv("https://rmorsomme.github.io/website/projects/training_set.csv")
```

and then perhaps fitting a model.

```{r}
m <- lm(mother_bmi ~ mother_weight_prepregnancy, data = d)
m
```

This is simply an illustration. In the project, you should not use `mother_bmi` as the response!

For the project, simply save the model you have selected in a RDATA file and send it to the instruction team.

```{r}
save(m, file = "my_predictive_model.RDATA")
rm(m) # removes the object `m` from the environment
```

Once we have everybody's model, we will use the following commands to load them into `R` and test them on new data.

```{r}
load("my_predictive_model.RDATA")

new_data <- tibble(mother_weight_prepregnancy = 150) # new data
predict(m, new_data)
```

## We can use RMD to write mathematical expressions

To write a small mathematical expression, simply write it between `$` signs.

For instance, $x = 5 + 9$.

To write longer expression, use two `$` signs.

$$
Y \approx \beta_0 + \beta_1 X_1 + \beta_2 X_2 
$$

To write a word in a math equation, use `\text{}`

$$
\text{BMI} \approx \beta_0 + \beta_1 \text{weight} + \beta_3 X^2
$$

## Running `R` code

You can run `R` code in `R` chunks:

```{r}
5+5
```

You can also directly run `R` code in a paragraph as follows: using \` r 5+5\`. For instance, the $R^2$ value of model `m` is `r round(glance(m)$r.squared, 3)`.

## Learning more about RMD

Come to OH, ask questions during/after class.

Check the RMD [cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/rmarkdown.pdf) or the longer [reference guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf).

## Overall goal of the prediction project

-   two baseline models (simple and full) + 3 models that you construct

-   feature engineering + model selection

-   the outcome of that procedure should be a model that predicts the response variable *reasonably* well

-   please refer to the rubric

## Missing values in the penguin dataset?

```{r}
d <- palmerpenguins::penguins

filter(d,  is.na(body_mass_g)) # keeps the rows with a missing value for body_mass_g
filter(d, !is.na(body_mass_g)) # keeps the rows without a missing value for body_mass_g
```

## Conclusion

This is the end of the main body

## Appendix

Figures should go here.

```{r}
ggplot(mpg, aes(cty, hwy)) + geom_point()
```
