---
title: "Homework 1 - Introduction to data"
author: "Author Name"
date: "`r Sys.Date()`"
output: openintro::lab_report
---


# Problem set - Chapters 1-2

## Exercise 1.2

## Exercise 1.8

## Exercise 1.12

## Exercise 1.20

## Exercise 2.8

## Exercise 2.12

## Exercise 2.14

## Exercise 2.22

## Exercise 2.22 - follow-up

-   Is your study blind?
-   Is it double-bind?
-   If it is not blind, can you make it blind?

\newpage

# Lab - Dr. Arbuthnot’s Baptism Records


```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Exercise 1

```{r view-girls-counts}
arbuthnot$girls
```

### Exercise 2

Insert any text here.

```{r trend-girls}
# Insert code for Exercise 2 here
```

### Exercise 3

Insert any text here.

```{r plot-prop-boys-arbuthnot}
# Insert code for Exercise 3 here
```

### Exercise 4

Insert any text here.

```{r dim-present}
# Insert code for Exercise 4 here
```

### Exercise 5

Insert any text here.

```{r count-compare}
# Insert code for Exercise 5 here
```

### Exercise 6

Insert any text here.

```{r plot-prop-boys-present}
# Insert code for Exercise 6 here
```

### Exercise 7

Insert any text here.

```{r find-max-total}
# Insert code for Exercise 7 here
```
