---
title: "HW 3 - Introduction to data"
subtitle: "Due Sunday, May 22, 9:00pm on Gradescope"
---

This assignment needs to be completed with RMarkdown and submitted as a PDF on [Gradescope](https://www.gradescope.com/courses/394638). Feel free to re-use the template provided for HW1.

When submitting your work on Gradescope, please assign a page for each question.

## Problem set (25 points)

-   7.6, a-c (3 points)
-   7.18 (2 points)
-   7.20 (4 points)
-   7.22 (7 points)
    -   fit the model in `R`; the variables `sho_gi` and `hgt` respectively correspond to shoulder girth and height.

        ```{r}
        d <- openintro::bdims
        ```

    -   do parts a-f
-   7.24 (5 points)
    -   fit the model in `R`

        ```{r}
        d <- MASS::cats
        ```

    -   do parts a-d
-   7.26 (3 points)
-   7.28 (1 point)

\newpage

## Lab exercises - the Human Freedom Index (25 points)

You can find the lab [here](http://openintrostat.github.io/oilabs-tidy/08_simple_regression/simple_regression.html)

-   Exercise 1 (2 points)
-   Exercise 2 (2 points)
-   Exercise 3 (3 points)
-   Exercise 4 (2 points)
-   Exercise 5 (2 points)
-   Exercise 6 (3 points)
-   Exercise 7 (2 points)
-   Exercise 8 (2 points)

Skip exercises 9-10.

-   Practice 1 (2 points)
-   Practice 2 (2 points)
-   Practice 3 (1 points) -- only evaluate linearity
-   Practice 4 (2 points)
