---
title: "HW 1 - Introduction to data"
subtitle: "Due Sunday, June 15, 9:00pm on Gradescope"
author: "Author Name"
date: "`r Sys.Date()`"
---

# Problem set (25 points)

## Exercise 1.2 (2 points)

## Exercise 1.8 (5 points)

## Exercise 1.12 (5 points)

## Exercise 1.20 (2 points)

## Exercise 2.8 (2 points)

## Exercise 2.12 (3 points)

## Exercise 2.14 (1 point)

## Exercise 2.22 (2 points)

## Exercise 2.22 - follow-up (3 points)

-   Is your study blind?
-   Is it double-bind?
-   If it is not blind, can you make it blind?

\newpage

# Lab exercises - Dr. Arbuthnot’s baptism records (25 points)

You can find the lab [here](http://openintrostat.github.io/oilabs-tidy/01_intro_to_r/intro_to_r.html)

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### Exercise 1 (2 points)

Insert any text here.

```{r view-girls-counts}
# Insert code for Exercise 1 here
```

### Exercise 2 (3 points)

Insert any text here.

```{r trend-girls}
# Insert code for Exercise 2 here
```

### Exercise 3 (4 points)

Insert any text here.

```{r plot-prop-boys-arbuthnot}
# Insert code for Exercise 3 here
```

### Exercise 4 (4 points)

Insert any text here.

```{r dim-present}
# Insert code for Exercise 4 here
```

### Exercise 5 (2 points)

Insert any text here.

```{r count-compare}
# Insert code for Exercise 5 here
```

### Exercise 6 (5 points)

Insert any text here.

```{r plot-prop-boys-present}
# Insert code for Exercise 6 here
```

### Exercise 7 (5 points)

Insert any text here.

```{r find-max-total}
# Insert code for Exercise 7 here
```
